Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms\thanks{This footnote leaves room for a piece of text saying how grateful we are to our funding agency. This includes a magic grant number, so it takes up three precious lines of space.}

### 2018-01-31 paper/paper.tex
Some of Patrick's changes

For the variable-ordering heuristic on \siplineref{variableordering} of \cref{algorithm:sip}, we use
smallest domain first, tiebreaking on highest degree. Variable ordering is not the focus of this
paper---as far as possible, we will leave it intact, so as not to affect the performance of the
algorithm on unsatisfiable instances.  For value ordering, which occurs on
\siplineref{valueordering} of \cref{algorithm:sip}, we will initially select vertices from highest
degree to lowest degree; we return to value ordering in the second half of this paper. Note that the
variable-ordering heuristic affects both unsatisfiable and satisfiable instances, whilst the
value-ordering heuristic is relevant only for satisfiable instances.

@@

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Obviously, such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\citep{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in the \emph{value}-ordering heuristic, and keep the
variable ordering fixed.\footnote{It may be possible to further improve \cref{algorithm:sip} by also
introducing randomness or some form of learning into its variable-ordering heuristic. However,
simultaneously introducing a second change would considerably complicate the empirical analysis.}
Using restarts on value-ordering heuristics appears to be uncommon, although
\citet{DBLP:conf/flairs/RazgonOP07} look at learning value-ordering heuristics from restarts, and
\citet{DBLP:conf/cp/ChuSS09} use a similar scheme in the context of parallel search.

@@

We use the Luby scheme \citep{DBLP:journals/ipl/LubySZ93} to determine when to restart. We count the
number of backtracks (that is, when we reach the end of the main $\KwSty{foreach}$ loop on
\siplineref{countrestarts}) to decide when to restart, rather than counting recursive calls.
Following established wisdom, we multiply each item in the Luby sequence by a magic constant---we
used SMAC \citep{DBLP:conf/lion/HutterHL11} to select the value 660. (We also tried a geometric
restart scheme. The results were less favourable---it appears that value-ordering heuristics benefit
from a very aggressive restart strategy.)

@@

In \cref{figure:restarts} we show the effects of adding restarts to the algorithm. With restarts,
the random value-ordering heuristic comfortably beats the degree strategy without restarts. In other
words, although having a good variable-ordering heuristic is beneficial, introducing randomness into
the search is better, if it is done alongside a mechanism to avoid heavy commitment to any
particular random choice. \Cref{figure:scatter-random} looks at this in more detail. The performance
on unsatisfiable instances, drawn as dark dots, remains unchanged as expected (which shows that the
overheads of nogoods and restarting are negligible). However, although better in aggregate on
satisfiable instances, on a case by case basis many instances perform substantially worse with
randomisation and restarts.

@@

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-random.pdf}
    \caption{Comparing basic backtracking with the degree heuristic, versus the Random heuristic
    with restarts.}
    \label{figure:scatter-random}
\end{figure}

@@

Having looked at subgraph isomorphism in detail, we now briefly discuss the more general maximum
common induced subgraph problem.  Two recent algorithms for this problem also make use of
backtracking search with degree as a value-ordering heuristic. The k${\downarrow}$ algorithm
\citep{DBLP:conf/aaai/HoffmannMR17} attempts to solve the problem by first trying to solve the
induced subgraph isomorphism problem, and then if that fails, retries allowing a single unmatched
vertex (and thus using weaker invariants), and so on. Due to its similarity to \cref{algorithm:sip},
we can introduce the same bias and restart strategy, requiring only a small amount of jiggery-pokery
to handle wildcards in nogoods.  Meanwhile, the McSplit${\downarrow}$ algorithm
\citep{DBLP:conf/ijcai/McCreeshPT17} uses a constraint programming style search, but with special
propagators and backtrackable data structures that exploit special properties of the problem. The
unconventional domain store used by McSplit${\downarrow}$ precludes the use of arbitrary unit
propagation, and so when introducing restarts, we cannot propagate using nogoods.  Instead, we can
only detect when we are inside an already-visited branch.  We must therefore use the one watched
literal scheme instead, and we also introduce a basic subsumption scheme to prune redundant clauses.

>>>>

For the variable-ordering heuristic on \siplineref{variableordering} of \cref{algorithm:sip}, we use
smallest domain first, tiebreaking on highest degree. Variable ordering is not the focus of this
paper---as far as possible, we use this same dynamic ordering throughout the paper, so as not to
affect the performance of the algorithm on unsatisfiable instances.  For value ordering, which
occurs on \siplineref{valueordering} of \cref{algorithm:sip}, we will initially select vertices from
highest degree to lowest degree; we return to value ordering in the second half of this paper. Note
that the variable-ordering heuristic affects both unsatisfiable and satisfiable instances, whilst
the value-ordering heuristic is relevant only for satisfiable instances.

@@

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Obviously, such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\citep{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in the \emph{value}-ordering heuristic, and continue
to use smallest domain first with static tiebreaking for variable-ordering.\footnote{It may be
possible to further improve \cref{algorithm:sip} by also introducing randomness or some form of
learning into its variable-ordering heuristic. However, simultaneously introducing a second change
would considerably complicate the empirical analysis.} Using restarts on value-ordering heuristics
appears to be uncommon, although \citet{DBLP:conf/flairs/RazgonOP07} look at learning value-ordering
heuristics from restarts, and \citet{DBLP:conf/cp/ChuSS09} use a similar scheme in the context of
parallel search.

@@

We use the Luby scheme \citep{DBLP:journals/ipl/LubySZ93} to determine when to restart. We count the
number of backtracks (that is, when we reach the end of the main $\KwSty{foreach}$ loop on
\siplineref{countrestarts}) to decide when to restart, rather than counting recursive calls.
Following established wisdom, we multiply each item in the Luby sequence by a magic constant---we
used SMAC \citep{DBLP:conf/lion/HutterHL11} to select the value 660. (We also tried a geometric
restart scheme, which gave less favourable results.)

@@

In \cref{figure:restarts} we show the effects of adding restarts to the algorithm. With restarts,
the random value-ordering heuristic comfortably beats the degree strategy without restarts. In other
words, although having a good value-ordering heuristic is beneficial, introducing randomness into
the search is better, if it is done alongside a mechanism to avoid heavy commitment to any
particular random choice. \Cref{figure:scatter-random} looks at this in more detail. The performance
on unsatisfiable instances, drawn as dark dots, remains unchanged as expected (which shows that the
overheads of nogoods and restarting are negligible). However, although better in aggregate on
satisfiable instances, on a case by case basis many instances perform substantially worse with
randomisation and restarts.

@@

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-random.pdf}
    \caption{Comparing basic backtracking with the degree heuristic, versus the random heuristic
    with restarts.}
    \label{figure:scatter-random}
\end{figure}

@@

Having looked at subgraph isomorphism in detail, we now briefly discuss the more general maximum
common induced subgraph problem.  Two recent algorithms for this problem also make use of
backtracking search with degree as a value-ordering heuristic. The k${\downarrow}$ algorithm
\citep{DBLP:conf/aaai/HoffmannMR17} attempts to solve the problem by first trying to solve the
induced subgraph isomorphism problem, and then if that fails, retries allowing a single unmatched
vertex (and thus using weaker invariants), and so on. Due to its similarity to \cref{algorithm:sip},
we can introduce the same bias and restart strategy, requiring only small changes
to handle wildcards in nogoods.

Meanwhile, the McSplit${\downarrow}$ algorithm
\citep{DBLP:conf/ijcai/McCreeshPT17} uses a constraint programming style search, but with special
propagators and backtrackable data structures that exploit special properties of the problem. The
unconventional domain store used by McSplit${\downarrow}$ precludes the use of arbitrary unit
propagation, and so when introducing restarts, we cannot propagate using nogoods.  Instead, we can
only detect when we are inside an already-visited branch.  We must therefore use the one watched
literal scheme instead, and we also introduce a basic subsumption scheme to prune redundant clauses.
